Controlling the dewetting morphologies of thin liquid films by switchable substrates

### 2023-11-14 Reply_Referees-2nd.tex

\definecolor{pyblue}{HTML}{1F77B4}
\definecolor{pyorange}{HTML}{FF7F0C}
\definecolor{pygreen}{HTML}{2CA02C}
\definecolor{pyred}{HTML}{D62728}

\newcommand{\JH}[1]{\textcolor{blue}{JH: #1}}
\begin{document}

% The referee PDF has been generated from r239 -- 'make referee' command.

\section*{Reply to 2nd report of Referee A}
\begin{itemize}  
\item[ \textbf{\underline{Comment 1.}}]
{ 
With the response given to me by the authors, I think that the
research is probably valid. I still fail to see why I should be very
excited about the findings, so I think that the manuscript would be
better suited for a more specialized journal, like Physical Review E.
The reason for this recommendation is that I can see what the
patterned substrate does, in the sense that the wavelength competes
with that of dewetting. That droplets can mode (move?) in a wettability
gradient has also been clear for a long time. I am wondering what is
there to be gained by combining these ideas, and I find the rationale
of the manuscript to be not extremely compelling. In my view, this
would be needed for PRL. I am sorry that I cannot be more positive.
}

\item[ \textbf{{Answer}}]
{
We thank the Referee for taking his/her time to read our 
manuscript once more and his/her subjective opinion. It is good to read 
that he/she thinks that our research is valid and that he/she finds the 
manuscript compelling to some extent.

We would like to stress once more that 
the main focus of the paper is not dewetting or droplet movement in a {\it stationary} wettability gradient, but dewetting 
morphologies over a {\it switchable substrate}, i.e. whose wettability pattern 
is {\it time-dependent}, whence the title 'Controlling dewetting morphologies of 
thin liquid films by switchable substrates'.

In this context, we have discovered a new morphological transition which we have fully characterized 
and theoretically interpreted in terms of the control parameter $\Gamma$.
This said, even in the static pattern case, although the dewetting problem was previously studied, we 
reported for the first time, together with a theoretical explanation, the absolutely non-trivial dependence of the rupture times on the pattern wavelength.
Let us remark again that a precise control of the liquid-vapour interface during a dynamic process is a highly desired feature for many applications. We did show that the liquid-vapour 
interface even after dewetting can be increased significantly due to
the discussed morphological transition.

More in general wetting, dewetting and the dynamics of droplets on complex substrates 
are state-of-the-art topics of interest to a broad audience, both in experimental and 
theoretical fluid physics. 
We also believe that a deeper understanding of the dynamics of thin liquid films on switchable 
substrates, which this paper contributes to, is essential to push forward the current reach of microfluidic technology.
}
\end{itemize}

\section*{Reply to 2nd report of Referee B}
\begin{itemize}  
\item[ \textbf{\underline{Comment 1.}}]
{ 
The authors have responded convincingly to my concerns and addressed
them sufficiently in the revised manuscript. I have also read the
responses to the other reports and I feel that the authors have
addressed these recommendations adequately. In view of this, and my
original favorable assessment, I recommend this manuscript for
publication.

I have just two remarks that are related to concern 5 in my original
report:

1. Most importantly, the authors have added footnote 26, which I find
to be unnecessary, but notice that the curvature has the denominator
to the power of 3/2.

2. The film thickness-dependent surface tension referred to in my
report is not related to the trivial small gradient approximation, and
appears whether the slope is small or large. It is a second order
effect which decays according to d Pi(h) / dh, and could be relevant
for thin films and helpful to address the contact line problem, but
not likely to change the results qualitatively.
}

\item[ \textbf{{Answer}}]
{
We thank the Referee for clarifying his/her question and we apologize for 
our misunderstanding (we have removed the footnote and the comment on 
the small gradient expansion of the curvature). 
We agree that, when thermal fluctuations are relevant,
it is important to consider the renormalized (i.e., height dependent) surface tension,
especially when it comes to the initial growth of the instability. It is, on the other 
hand, reasonable to expect, as the Referee remarks, that in terms of the dewetting 
morphology (and its relation to the time-dependence of the substrate) the phenomenology
should not be particularly affected, at least qualitatively.
}
\end{itemize}

@@

Not only do we present our numerical and analytical results in this letter, but we also supply a potential realization for a lab experiment. 
We provide a schematic setup using a photoswitchable substrate and a digital multimirror device. 
As we are well aware that this is not sufficient for an experiment we further identify the relevant length and time scales for both the film and the apparatus.
% Hoping that experimentalist confirm transition we supply a way to construct pattern Eq. (3).
%We therefore have a hard time understanding why it seems impossible to the referee to construct the pattern equation %(3), as he/she does not give us a single increment of information in his/her comment.
}

\item[ \textbf{\underline{Comment 2.}}] 
{
This work is completed based on the lubrication equation for the thin film flows.
From theoretical/mathematical aspects, the novelty of this work, in my opinion, is not strong enough. 
This lubrication model employed here was proposed about 40 years ago and has been thoroughly investigated~\cite{RevModPhys.69.931, RevModPhys.81.1131}. 
There are no improvements or modifications to the model in this work. 
Additionally, the model for the rupture time comes from a simple scaling analysis, which is not very convincing when considering other classical mathematical approaches, such as linear instability analysis, similarity transformation, etc.
}

\item[ \textbf{{Answer}}]
{
We find it puzzling that the referee arrives at the conclusion that working on a theory which is 40 years old does not allow for the novelty attributed to PRL.
A good counter example is: M. Wilczek et al,
"Sliding drops – ensemble statistics from single drop bifurcations", {\it Phys. Rev. Lett.} {\bf 119}, 204501 (2017), in which Wilczek et al. use the thin film equation on an inclined plane.
Similar arguments can be constructed about the Navier-Stokes equation which is 180 years old, still there is a lot of research dedicated towards, e.g., turbulence.

We agree with the referee that from a purely mathematical point of view lubrication theory is untouched.
There is no reason or advantage to add artificial complexity to a well working theory which has been developed for this singular purpose.
However, when it comes to dynamical effects and non-linear behaviour of thin films, the topic this letter revolves around, mathematics falls short and only numerical simulation or experiments offer deeper insights, which we also outline in the first paragraph of the introduction:

``\textit{While a consistent body of theoretical/computational work was devoted to processes on static heterogeneous substrates, the time dependent case is still almost unexplored, with few exceptions focusing on single droplet spreading and sliding~\cite{GrawitterStark1,GrawitterStark2,ThieleHartmann} or limited to analysing the linear regime~\cite{suman2006dynamics}.}''

To this end we use a novel lattice Boltzmann method to solve the thin film equation at time scales where initial small perturbations have grown to similar sizes as $h_0$.
It is well known that the rupture time marks the end of the linear regime and linear instability theory as suggested by the referee, which has been used in our previous work~\cite{PhysRevE.104.034801}. 
However, it is simply not able to explain our observations, as $\tau_r$ is dependent on $\Gamma$. 
Observing the morphological transition during dewetting from a droplet to a rivulet state is by no means covered by linear analysis. 
It is a dynamic competition of relevant velocities which to the best knowledge of the authors has not been discussed or shown before.
}

### 2023-11-14 main_exciting.tex
Work on references and some minor editing.

\documentclass[twocolumn,amsmath,amssymb,showpacs,prl,superscriptaddress]{revtex4-1} %,preprint,jcp,nofootinbib

\bibstyle{apsrev}
\usepackage{bm}%
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{color}
\usepackage{xcolor}
\usepackage{physics}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}%
\expandafter\ifx\csname package@font\endcsname\relax\else
 \expandafter\expandafter
 \expandafter\usepackage
 \expandafter\expandafter
 \expandafter{\csname package@font\endcsname}%
\fi
\hyphenation{title}
\newcommand{\JH}[1]{\textcolor{blue}{ JH: #1}}
\newcommand{\REV}[1]{\textcolor{red}{#1}}
%\newcommand{\REV}[1]{#1}

@@

\begin{abstract}
\noindent Switchable and adaptive substrates emerged as valuable tools for controlling wetting and actuation of droplet motion. Here, we report a computational study of the dynamics of an unstable thin liquid film deposited on a switchable substrate, modeled with a space and time varying contact angle.
With a static pattern, all fluid is drained into droplets located around contact angle minima, whereas for a sufficiently large rate of wettability variation a state consisting of metastable rivulets is observed. 
A criterion discriminating whether rivulets occur or not is identified in terms of a single dimensionless parameter.
Finally, we show and explain theoretically how the film rupture times, droplet shape and rivulet life time depend on the pattern wavelength and speed.  
\end{abstract}

@@

\noindent {\it Introduction}. Wet surfaces and droplets are part of our every-day experience and of numerous industrial processes including coating, tribology, painting and printing, to name but a few~\cite{gross1980fluid,szeri2010fluid,doi:10.1146/annurev.fluid.31.1.347,DASILVASOBRINHO19991204,singh2010inkjet,Wijshoff2010}. 
Moreover, the continuously growing interest for lab-on-a-chip devices~\cite{C6LC00387G,Focke} as well as for printable electronics or printable photovoltaics~\cite{Luechinger_2008,RH22}, whose efficiency relies crucially on a precise control of material deposition upon (de-)wetting of liquid films, drew the attention to applications where the substrate is adaptive or switchable, i.e. it is not inert but responds dynamically to external stimuli or to the evolution of the coating liquid film itself~\cite{ButtEtAl_Langmuir2018}. 
Several realizations of switchable and adaptive substrates have been proposed~\cite{XiCSR2010}, involving smart materials such as polymer brushes~\cite{CohenStuartEtAl_NatMat2010}, thermal-responsive hydrogels~\cite{ChenEtAl_SM2010}, light-responsive molecules and microstructures~\cite{IchimuraEtAl_Science2000} or processes such as electrowetting~\cite{MugeleEtAl_JPCM2005}.
Probably the simplest, yet non-trivial, modelling of a switchable substrate can be realized by a space and time dependent wettability pattern~\cite{GrawitterStark1}. 
While a consistent body of theoretical/computational work was devoted to processes on static heterogeneous substrates, the time dependent case is still almost unexplored, with few exceptions focusing on single droplet spreading and sliding~\cite{GrawitterStark1,GrawitterStark2,ThieleHartmann} or limited to analysing the linear regime~\cite{suman2006dynamics}.

Here, we study by means of numerical simulations the full dewetting dynamics of a thin liquid film deposited on a substrate with a time varying wettability pattern, from rupture to the long time morphology. 
We identify two regimes where the rupture times grow with the pattern wavelength either linearly (on a static pattern) or attain a constant value (in the time dependent case), for short wavelengths, and approach a quadratic law as the wavelength increases. These observations are then explained theoretically.
We show that, by tuning the rate of change (the ``speed'') of the underlying pattern, one can control, to some extent, the dewetting morphology. In particular, for large enough pattern speeds, we detect a metastable state, where the film retracts into metastable rivulets, eventually breaking up into multiple droplets.

@@

\noindent {\it Method.} In order to simulate the dewetting dynamics on patterned, ``switchable'', substrates, we integrate numerically the thin-film equation~\cite{RevModPhys.69.931,RevModPhys.81.1131} 

@@

by means of a recently developed lattice Boltzmann (LB) scheme~\cite{PhysRevE.100.033313,PhysRevE.104.034801, Zitz2022}.

@@

The film pressure $p(\mathbf{x},t)$ consists of the sum of the Laplace and disjoining pressures, that is $p(\mathbf{x},t) = -\gamma \nabla^2 h - \Pi$ \REV{($\gamma$ is the surface tension)}~\footnote{\REV{The Laplace pressure is the product of the surface tension 
and the local curvature. Therefore, its full expression reads $p_L = -\gamma \nabla^2 h/\sqrt{1+|\nabla h|^2}$. 
In the lubrication regime, whereby $|\nabla h| \ll 1$, the denominator can be safely neglected 
and the pressure reduces to the form involving only the Laplacian of the thickness 
field}~\cite{Benet2014,juanes2018}}.

@@

As such, it contains the information on the liquid/solid and solid/gas interactions and, hence, on the wettability, which is parameterized in terms of the contact angle $\theta$~\cite{RevModPhys.81.739, SCHWARTZ1998173}. 

@@

\REV{\Pi(h,\theta) = \frac{2\gamma}{h_{\ast}}(1-\cos(\theta(\mathbf{x},t)))
  f\left(\frac{h}{h_{\ast}}\right),}

@@

\REV{where $f(\xi)=\xi^{-3} - \xi^{-2}$}. $h_{\ast}$ is the height at which $\Pi$ vanishes and sets the precursor layer thickness~\cite{SuppMat}.

@@

\REV{We decorate the substrate with a checkerboard pattern, a common choice that generalizes the broken homogeneity of the stripes to two 
directions~\cite{Jalali2018,Nagayama2020,Das2020}.}

@@

% \REV{A possible laboratory experiment for the here mentioned approach consists of a light switchable substrate with a thin film cast on it~\cite{becker2003complex, IchimuraEtAl_Science2000}. 
% The external stimulus, our moving pattern, can be realized, e.g., by using a digital multimirror device (DMD)~\cite{doi:10.1021/jp301092y,doi:10.1126/science.aax8760}.}

@@

\REV{Lengths and time scales will be expressed, respectively, in units of the mean film height, $h_0$ (which is constant in time, due to mass conservation), and of $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}$, the inverse growth rate of the most unstable mode, whose wavenumber is $q_0$, of a spinodally dewetting film~\cite{Mecke_2005}. On a uniform substrate, with constant contact angle $\theta^{(u)}$, the wavenumber reads 
$(q^{(u)}_0)^2 = h_{\ast}^{-2}(1-\cos \theta^{(u)})f^{\prime}(h_0/h_{\ast})$~\cite{Mecke_2005,PhysRevE.100.023108}. In our patterned case, we define 

@@

$\lambda_s = 2\pi/q_0$.}

@@

Fig.~\ref{fig:handtheta} shows $h(\mathbf{x},t)$ (droplets) and $\theta(\mathbf{x})$ (color coded) for $v_{\theta} = 0$ (i.e. the static case) and \REV{$\lambda = 256 h_0$}~\footnote{For a better visualization double the domain length $L$ and periodically continue the image.}, in late stages of dewetting.

@@

It is conveyed that, overall, rupture occurs earlier on the static substrate, suggesting that the time variation tends to stabilize the film, in agreement with linear stability analysis results~\cite{suman2006dynamics}.

@@

\begin{multline}\label{eq:taur_l2}
    \tau_r \sim t_{\theta} \sim  \delta h/\dot{(\delta h)} \propto \frac{3\mu}{h_0^3}(\partial_x^2 (\partial_h\Pi (h_0)))^{-1} \sim \\
    t_0 \left(\frac{q_{\theta}}{q_0}\right)^{-2} \propto t_0 q_0^2 \lambda^2.
\end{multline}

@@

 %\tau_r \sim \tau_R \propto \frac{1}{U_{\Theta}}\lambda,

@@

where $U_{\Theta}$ is the retraction speed $U_{\Theta} = \frac{\gamma \Theta^3}{9\mu}$~\cite{Edwards2016}, with $\Theta = \max_{\mathbf{x}}\theta(\mathbf{x})$.

@@

    % The dashed lines indicate the geometrically expected values of the droplet height $h_d$, assuming monodispersity and a perfect spherical cap shape. 

@@

      which is $2\left(\frac{L}{\lambda}\right)^2$. The napshots depict the stationary droplet states as grey-scale images

@@

% the clusters are determined by means of a Hoshen-Kopelman algorithm~\cite{HK}.

@@

\REV{In the main panel of the figure, the height fluctuations $\Delta h(t) = \max_{\mathbf{x}}\{h(\mathbf{x},t)\}-\min_{\mathbf{x}}\{h(\mathbf{x},t)\}$
grow in time until film rupture and then settle to a constant value. This represents a measure of the mean droplet height $h_d$ (since
droplets are essentially monodisperse), decreasing with the pattern wavelength (as expected, due to a decreasing droplet volume,
$V_d = \frac{h_0 \lambda^2}{2}$).}\\
%In the steady state, since the droplets are essentially monodisperse, this observable . Assuming that the droplet shape is a spherical cap,
%it can be estimated as $h_d =\left(\frac{3V_d(1-\cos(\theta_d))}{\pi(2+\cos(\theta_d))}\right)^{1/3}$, where  is the droplet volume and
%$\theta_d$ is the local contact angle (i.e. the value at the contact line). This, in turn, depends on $h_d$,
%through Eq.~(\ref{eq:sinetheta}), thus making the one above an implicit equation to be solved numerically for $h_d = h_d(\lambda)$. 
%$\Delta h(t)$ is reported in Fig.~\ref{fig:clusters_v0_sine} for three different wavelengths, 
%$\lambda/h_0 = \{170,256,512 \}$, together with the theoretically expected $h_d(\lambda)$ (depicted with lines),
%showing excellent agreement.\\

@@

For $v_{\theta} = 1.7 v_0$ we still observe the formation of droplets, similarly to the stationary case ($v_{\theta} = 0$). However, these are transported with the contact angle minima, reproducing a somehow similar behaviour recently described in a numerical study of a droplet on a moving wettability step~\cite{GrawitterStark1}.

@@

In order to better characterize the various morphologies we apply the theory of Minkowski's functionals. 
In particular, we employ the second order Minkowski structure metric, $q_2$~\cite{doi:10.1063/1.4774084, Schaller2020}, which can be computed from a Voronoi tessellation of the set of discrete points $(x_i, y_i)$ on the 2D lattice, such that the height field lies above a certain threshold\footnote{The expression is: $q_2 = \frac{1}{N}\sum_j \frac{1}{P_j}|\sum_k L^{(j)}_k e^{2i\phi^{(j)}_k}|$, where the inner sum runs over the edges of length $L^{(j)}_k$, of the $j$-th Voronoi cell, whose perimeter is $P_j$, and $\phi^{(j)}_k$ is the polar angle of the normal to the $k$-th edge. 
The outer sum represents an ensemble average over the $N$ points in the set.}. The $q_2$ metric quantifies the degree of anisotropy of the dewetting morphology, so it takes relatively large values if the structures formed display a preferential direction. 
Measuring $q_2$ then enables us to clearly distinguish between the formation of droplets and rivulets: much larger $q_2$ values are attained for the latter type of structure, as we can see in Fig.~\ref{fig:msm_q2}.
%Our simulations with $\lambda = L/2$ show a clear difference between $v_{\theta} \leq 0.1v_0$ and $v_{\theta}>0.1v_0$  in $q_2$.
We observe, on the other hand, that such rivulets are metastable and eventually break up into droplets, as indicated by the collapse of $q_2$ at later times. 
Notice, though, that the $q_2$ signal for any $v_{\theta} >0$ always stays above the one for the static case, suggesting that even the smallest pattern velocity introduces a sizeable deformation of the spherical cap shape.
\begin{figure}
    \centering
    % \includegraphics[width=0.45\textwidth]{Figures/Lig_stab_inset_delta_h.pdf}
    \includegraphics[width=0.4\textwidth]{Figure_5.pdf}
    \caption{MAIN PANEL: Rivulet life-times, $\tau_{\text{riv}}$, for various $\Gamma$.
    The dashed line is a guide to the eye to highlight the logarithmic dependence, in agreement with 
    the theoretical prediction, Eq.~(\ref{eq:rivlt}).
    INSET: Height fluctuations, $\Delta h(t)$, vs time, along the rivulet axis, for three different $\Gamma$.
    }
    \label{fig:stab_ligs_lam2}
\end{figure}
The breakup is the result of a varicose mode of the rivulet~\cite{doi:10.1063/1.3211248, PhysRevE.77.061605}, whose wavelength is $\approx\lambda$, such that only $N_{\infty}/2$ droplets are counted after breakup. 
These droplets show a peculiar dynamics, characterized by a periodic sequence of spreading and retraction, driven by the pattern, that we dub "pumping state"~\footnote{see movie ligament\_formation\_and\_breakup.mp4 in~\cite{SuppMat}.}.

@@

\REV{If $U_{\Theta}$ is large as compared to $v_{\theta}$, the film retraction is faster than 
the local contact angle variation and thus droplets form.}
%the transport due to the contact angle field and thus droplets from. 
However, if $v_{\theta}$ is larger than $U_{\Theta}$, then the retracting film has too little time to form droplets and ends up in the metastable rivulet state. \REV{It appears, therefore, natural to consider the ratio of these two velocities, 
$\Gamma \equiv v_{\theta}/U_{\theta}$, as the discriminating parameter.}
%Recalling the expression (\ref{eq:t0}) for the reference velocity $v_0$, %we define the parameter 
%\begin{equation}\label{eq:vel_ratio}
%    \Gamma = \frac{v_{\theta}}{U_{\Theta}} = \frac{3\lambda h_0^3 %q_0^4}{\Theta^3}\chi 
%\end{equation}
%as the ratio of these two velocities,
%where $\chi \equiv v_{\theta}/v_0$. 

@@

The rivulet life-time can be seen as the rupture time of the structure along its axis, hence such that $\Delta h (\tau_{\text{riv}}) \sim h_0$~\cite{PhysRevE.104.034801}, which yields

@@

\REV{We envisage a possible realization of a 
dewetting experiment on a switchable substrate of the type modeled by the spatio-temporal 
contact angle (\ref{eq:sinetheta}). One may think of a thin liquid film cast on a light responsive substrate~\cite{IchimuraEtAl_Science2000}, under the action of controlled 
external stimuli (a light emitter). 

@@

This technology was effectively used for thin film experiments and additive manufacturing~\cite{doi:10.1021/jp301092y, doi:10.1126/science.aax8760}. It allows for fast temporal modulations of the optical signal (with frequencies up to $\approx 16 \, \text{kHz}$)
with spatial resolution of $\approx 10 \times 10 \, \mu \text{m}^2$ (the size of a pixel).
Considering as a reference, for instance, the system studied in~\cite{becker2003complex,PhysRevLett.99.114503}, namely a $\sim 4 \, \text{nm}$ thick film 
of polystyrene deposited on an oxidized silicon wafer,  
we evaluate the retraction speed 
$U_{\Theta} = \frac{\Theta^3 \gamma}{9 \mu}$ to be 
$U_{\Theta} \approx 10^{-2} \, \mu \text{m}/\text{s}$. 
The (minimum) {\it pattern speed} can be 
estimated from the pixel size with frame rate $\sim 1 s^{-1}$ as 

>>>>

\documentclass[amsmath,amssymb,showpacs,prf,superscriptaddress, longbibliography]{revtex4-1} %,preprint,jcp,nofootinbib
%\documentclass[twocolumn,amsmath,amssymb,showpacs,prl,superscriptaddress]{revtex4-1} %,preprint,jcp,nofootinbib

\bibstyle{apsrev}
\usepackage{bm}%
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{color}
\usepackage{xcolor}
\usepackage[normalem]{ulem}
\usepackage{physics}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}%
\expandafter\ifx\csname package@font\endcsname\relax\else
 \expandafter\expandafter
 \expandafter\usepackage
 \expandafter\expandafter
 \expandafter{\csname package@font\endcsname}%
\fi
\hyphenation{title}
\newcommand{\JH}[1]{\textcolor{blue}{ JH: #1}}
\newcommand{\REV}[1]{\textcolor{red}{#1}}
%\newcommand{\REV}[1]{#1}

@@

\begin{abstract}
 \noindent Switchable and adaptive substrates emerged as valuable tools for controlling wetting and actuation of droplet motion. 
 Here, we report a computational study of the dynamics of an unstable thin liquid film deposited on a switchable substrate, modelled with a space and time varying contact angle.
 \REV{For a sufficiently large rate of wettability variation a novel topological transition appears. 
 Instead of breakup into droplets, as expected for a substrate with multiple wetting minima, a metastable rivulet state emerges}. 
 A criterion discriminating whether rivulets occur or not is identified in terms of a single dimensionless parameter.
 Finally, we show and \REV{derive} theoretically how the film rupture times, droplet shape and rivulet life time depend on the pattern wavelength and speed.
\end{abstract}

@@

\noindent {\it Introduction}. Wet surfaces and droplets are part of our every-day experience and of numerous industrial processes including coating, tribology, painting and printing, to name but a few~\cite{grossFluidFilmLubrication1980a,szeriFluidFilmLubrication2010,quereFluidCoatingFiber1999,dasilvasobrinhoStudyDefectsUltrathin1999,singhInkjetPrintingProcess2010,wijshoffDynamicsPiezoInkjet2010}. 
Moreover, the continuously growing interest for lab-on-a-chip devices~\cite{samieiReviewDigitalMicrofluidics2016,fockeLabonaFoilMicrofluidicsThin2010} as well as for printable electronics or printable photovoltaics~\cite{luechingerGraphenestabilizedCopperNanoparticles2008,ronsinFormationCrystallineBulk2022}, whose efficiency relies crucially on a precise control of material deposition upon (de-)wetting of liquid films, drew the attention to applications where the substrate is adaptive or switchable, i.e. it is not inert but responds dynamically to external stimuli or to the evolution of the coating liquid film itself~\cite{buttAdaptiveWettingAdaptation2018}. 
Several realizations of switchable and adaptive substrates have been proposed~\cite{xinReversiblySwitchableWettability2010}, involving smart materials such as polymer brushes~\cite{stuartEmergingApplicationsStimuliresponsive2010}, thermal-responsive hydrogels~\cite{chenThermalresponsiveHydrogelSurface2010}, light-responsive molecules and microstructures~\cite{ichimuraLightDrivenMotionLiquids2000} or processes such as electrowetting~\cite{mugeleElectrowettingConvenientWay2005}.
\REV{Modelling the coupled problem of a thin film on a switchable substrate and as such the dynamics of the substarte as well as the film is far from being trivial but can be realized by a space and time dependent wettability pattern~\cite{grawitterSteeringDropletsSubstrates2021}}. 
While a consistent body of theoretical/computational work was devoted to processes on static heterogeneous substrates, the time dependent case is still almost unexplored, with few exceptions focusing on single droplet spreading and sliding~\cite{grawitterSteeringDropletsSubstrates2021,grawitterDropletsSubstratesOscillating2021,thieleGradientDynamicsModel2020} or limited to analysing the linear regime~\cite{sumanDynamicsThinLiquid2006}.

Here, we study by means of numerical simulations the full dewetting dynamics of a thin liquid film deposited on a substrate with a time varying wettability pattern, starting from \REV{a slightly perturbed film to film} rupture \REV{and} long time morphology. 
We identify two regimes where the rupture times grow with the pattern wavelength either linearly (on a static pattern) or attain a constant value (in the time dependent case), for short wavelengths, and approach a quadratic law as the wavelength increases. 
These observations are then explained theoretically. 
\REV{We show that, by tuning the rate of change of the underlying pattern, one can control the dewetting morphology, which is often desired in microfluidic applications.} 
In particular, for large enough pattern speeds, we detect a \REV{novel} state, where the film retracts into metastable rivulets, eventually breaking up into multiple droplets.

@@

\noindent {\it Method.} In order to simulate the dewetting dynamics on patterned, ``switchable'', substrates, we integrate numerically the thin-film equation~\cite{oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009} 

@@

by means of a recently developed lattice Boltzmann (LB) scheme~\cite{PhysRevE.100.033313,zitzLatticeBoltzmannSimulations2021, zitzSwalbeJlLattice2022a}.

@@

The film pressure $p(\mathbf{x},t)$ consists of the sum of the Laplace and disjoining pressures, that is $p(\mathbf{x},t) = -\gamma \nabla^2 h - \Pi$ ($\gamma$ is the surface tension)

@@

As such, it contains the information on the liquid/solid and solid/gas interactions and, hence, on the wettability, which is parameterized in terms of the contact angle $\theta$~\cite{bonnWettingSpreading2009, schwartzSimulationDropletMotion1998}. 

@@

\Pi(h,\theta) = \frac{2\gamma}{h_{\ast}}(1-\cos(\theta(\mathbf{x},t)))
  f\left(\frac{h}{h_{\ast}}\right),

@@

where $f(\xi)=\xi^{-3} - \xi^{-2}$. $h_{\ast}$ is the height at which $\Pi$ vanishes and sets the precursor layer thickness~\cite{SuppMat}.

@@

We decorate the substrate with a checkerboard pattern, a common choice that generalizes the broken homogeneity of the stripes to two 
directions~\cite{jalaliFabricationCharacterizationScalable2018,nagayamaIntermediateWettingState2020,dasSurfacedirectedSpinodalDecomposition2020}.

@@

@@

Lengths and time scales will be expressed, respectively, in units of the mean film height, $h_0$ (which is constant in time, due to mass conservation), and of $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}$, the inverse growth rate of the most unstable mode, whose wavenumber is $q_0$, of a spinodally dewetting film~\cite{meckeThermalFluctuationsThin2005}. 
On a uniform substrate, with constant contact angle $\theta^{(u)}$, the wavenumber reads $(q^{(u)}_0)^2 = h_{\ast}^{-2}(1-\cos \theta^{(u)})f^{\prime}(h_0/h_{\ast})$~\cite{meckeThermalFluctuationsThin2005,zhangMolecularSimulationThin2019}. 
In our patterned case, we define 

@@

$\lambda_s = 2\pi/q_0$.

@@

Fig.~\ref{fig:handtheta} shows $h(\mathbf{x},t)$ (droplets) and $\theta(\mathbf{x})$ (color coded) for $v_{\theta} = 0$ (i.e. the static case) and $\lambda = 256 h_0$~\footnote{For a better visualization double the domain length $L$ and periodically continue the image.}, in late stages of dewetting.

@@

It is conveyed that, overall, rupture occurs earlier on the static substrate, suggesting that the time variation tends to stabilize the film, in agreement with linear stability analysis results~\cite{sumanDynamicsThinLiquid2006}.

@@

\begin{equation}\label{eq:taur_l2}
    \tau_r \sim t_{\theta} \sim  \delta h/\dot{(\delta h)} \propto \frac{3\mu}{h_0^3}(\partial_x^2 (\partial_h\Pi (h_0)))^{-1} \sim t_0 \left(\frac{q_{\theta}}{q_0}\right)^{-2} \propto t_0 q_0^2 \lambda^2.

@@

@@

where $U_{\Theta}$ is the retraction speed $U_{\Theta} = \frac{\gamma \Theta^3}{9\mu}$~\cite{edwardsNotSpreadingReverse2016a}, with $\Theta = \max_{\mathbf{x}}\theta(\mathbf{x})$.

@@

@@

      which is $2\left(\frac{L}{\lambda}\right)^2$. The snapshots depict the stationary droplet states as grey-scale images

@@

@@

In the main panel of the figure, the height fluctuations $\Delta h(t) = \max_{\mathbf{x}}\{h(\mathbf{x},t)\}-\min_{\mathbf{x}}\{h(\mathbf{x},t)\}$ grow in time until film rupture and then settle to a constant value. 
This represents a measure of the mean droplet height $h_d$ (since droplets are essentially monodisperse), decreasing with the pattern wavelength (as expected, due to a decreasing droplet volume,
$V_d = \frac{h_0 \lambda^2}{2}$).\\

@@

For $v_{\theta} = 1.7 v_0$ we still observe the formation of droplets, similarly to the stationary case ($v_{\theta} = 0$). 
However, these are transported with the contact angle minima, reproducing a somehow similar behaviour recently described in a numerical study of a droplet on a moving wettability step~\cite{grawitterSteeringDropletsSubstrates2021}.

@@

In order to better characterize the various morphologies we apply the theory of Minkowski's functionals. 
In particular, we employ the second order Minkowski structure metric, $q_2$~\cite{mickelShortcomingsBondOrientational2013, schallerPapaya22DIrreducible2020}, which can be computed from a Voronoi tessellation of the set of discrete points $(x_i, y_i)$ on the 2D lattice, such that the height field lies above a certain threshold\footnote{The expression is: $q_2 = \frac{1}{N}\sum_j \frac{1}{P_j}|\sum_k L^{(j)}_k e^{2i\phi^{(j)}_k}|$, where the inner sum runs over the edges of length $L^{(j)}_k$, of the $j$-th Voronoi cell, whose perimeter is $P_j$, and $\phi^{(j)}_k$ is the polar angle of the normal to the $k$-th edge. 
The outer sum represents an ensemble average over the $N$ points in the set.}. The $q_2$ metric quantifies the degree of anisotropy of the dewetting morphology, so it takes relatively large values if the structures formed display a preferential direction. 
Measuring $q_2$ then enables us to clearly distinguish between the formation of droplets and rivulets: much larger $q_2$ values are attained for the latter type of structure, as we can see in Fig.~\ref{fig:msm_q2}.
We observe, on the other hand, that such rivulets are metastable and eventually break up into droplets, as indicated by the collapse of $q_2$ at later times. 
Notice, though, that the $q_2$ signal for any $v_{\theta} >0$ always stays above the one for the static case, suggesting that even the smallest pattern velocity introduces a sizeable deformation of the spherical cap shape.
\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{Figure_5.pdf}
    \caption{MAIN PANEL: Rivulet life-times, $\tau_{\text{riv}}$, for various $\Gamma$.
    The dashed line is a guide to the eye to highlight the logarithmic dependence, in agreement with 
    the theoretical prediction, Eq.~(\ref{eq:rivlt}).
    INSET: Height fluctuations, $\Delta h(t)$, vs time, along the rivulet axis, for three different $\Gamma$.
    }
    \label{fig:stab_ligs_lam2}
\end{figure}
The breakup is the result of a varicose mode of the rivulet~\cite{diezBreakupFluidRivulets2009a, mechkovStabilityLiquidRidges2008}, whose wavelength is $\approx\lambda$, such that only $N_{\infty}/2$ droplets are counted after breakup. 
These droplets show a peculiar dynamics, characterized by a periodic sequence of spreading and retraction, driven by the pattern, that we dub "pumping state"~\footnote{see movie ligament\_formation\_and\_breakup.mp4 in~\cite{SuppMat}.}.

@@

If $U_{\Theta}$ is large as compared to $v_{\theta}$, the film retraction is faster than the local contact angle variation and thus droplets form. 
However, if $v_{\theta}$ is larger than $U_{\Theta}$, then the retracting film has too little time to form droplets and ends up in the metastable rivulet state. 
It appears, therefore, natural to consider the ratio of these two velocities, $\Gamma \equiv v_{\theta}/U_{\theta}$, as the discriminating parameter.

@@

The rivulet life-time can be seen as the rupture time of the structure along its axis, hence such that $\Delta h (\tau_{\text{riv}}) \sim h_0$~\cite{zitzLatticeBoltzmannSimulations2021}, which yields

@@

We envisage a possible realization of a dewetting experiment on a switchable substrate of the type modelled by the spatio-temporal contact angle (\ref{eq:sinetheta}). 
One may think of a thin liquid film cast on a light responsive substrate~\cite{ichimuraLightDrivenMotionLiquids2000}, under the action of controlled external stimuli (a light emitter). 

@@

This technology was effectively used for thin film experiments and additive manufacturing~\cite{vieyrasalasActiveControlEvaporative2012, sahaScalableSubmicrometerAdditive2019}. 
It allows for fast temporal modulations of the optical signal (with frequencies up to $\approx 16 \, \text{kHz}$) with spatial resolution of $\approx 10 \times 10 \, \mu \text{m}^2$ (the size of a pixel).
Considering as a reference, for instance, the system studied in~\cite{beckerComplexDewettingScenarios2003,fetzerThermalNoiseInfluences2007}, namely a $\sim 4 \, \text{nm}$ thick film of polystyrene deposited on an oxidized silicon wafer, we evaluate the retraction speed $U_{\Theta} = \frac{\Theta^3 \gamma}{9 \mu}$ to be $U_{\Theta} \approx 10^{-2} \, \mu \text{m}/\text{s}$. 
The (minimum) {\it pattern speed} can be estimated from the pixel size with frame rate $\sim 1 s^{-1}$ as $v_{\theta} \sim 10 \, \mu \text{m}/\text{s}$, which would result in $\Gamma \sim 10^3$, i.e. well within the rivulet regime ($\Gamma > 1$). 
